---
title: "Lab 2 - Data Structures in R"
subtitle: "Environmental Data Analytics"
author: "John Fay & Luana Lima"
date: "Fall 2024"
---

## Objective
Discuss and navigate different data types in R
 - Vectors
 - Matrices
 - Arrays
 - Data frames

---

## Data Types in R
R treats objects differently based on their characteristics. For more information, please see: <https://www.statmethods.net/input/datatypes.html>.

-   **Vectors** 1 dimensional structure that contains elements of the same type.

-   **Matrices** 2 dimensional structure that contains elements of the same type.

-   **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

-   **Lists** Ordered collection of elements that can have different modes.

-   **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

---

## Vectors
### Creating Vectors
Add comments below that specify what kind of vectors are created. (First answer is provided.)

```{r Creating vectors}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # Vector type: Numeric - need to be all the same type
vector1
class(vector1)

vector2 <- c("one","two","three") # Vector type: character
vector2

vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) # Vector type: logic
vector3
```

### Vector Operations
Add comments below to indicate what each operation does.(First answer is provided.)

```{r Vector operatons}
# Returns the 3rd item in the vector
vector1[3]

#returns max value of vector 1
max(vector1)

#Create another vector of the values 10 thru 15
vector4 <- c(10:15)

#combining vectors 1 and 4 - as long as they are same data type
c(vector1, vector4)

#multiplying numbers in both vectors - same number of elements and multiplied based on position
vector1 * vector4

#all elements multiplied by 10
vector1 * 10

#lower to uppercase function 
toupper(vector2) 

#! denying the object - get the opposite of what you have vector so true becomes false
!vector3

```

### Exercise 1a: Vector
Find a 5-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create a vector representing the high temperature on each of the five days.
<https://www.wunderground.com/forecast/us/nc/durham/KNCDURHA138>

```{r Vector Exercise}

#Assign to the variable 'temps_F' a vector of high temperatures in F
temps_F <- c(80,82,80,75,74)

#What is the average of these temperatures 
mean_tempsF <- mean(temps_F)

#Compute the difference between each temperature and the average calculated above
temps_F - mean_tempsF

#Subtract 32 and multiply by 5/9ths to create a vector in Celsius
temps_C <- (temps_F - 32) *5/9

```

## Matrices
###Creating Matrices
Complete the comments below:
```{r Creating Matrices}
# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # number of rows and number of columns
matrix1
#matrix1 has __ rows and __ columns
#organized by column is default

matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #by row means organized by row
matrix2
#matrix2 has __ rows and __ columns


matrix3 <- matrix(
  data = 1:20, 
  nrow = 5, 
  ncol = 4, 
  byrow = TRUE, # return after comma continues the line
  dimnames = list(
    c("uno", "dos", "tres", "quatro", "cinco"),
    c("un", "deux", "trois", "quatre")
    )
  ) 

```

### Matrix Operations
Add comments in the code chunk below to indicate type of vectors created and what each operation does.
```{r Matrix Operations}

matrix1[4, 1] #row 4 column 1

matrix1[4,  ] #all the elements on row 4

matrix1[ , 3] #all elements on column 3

matrix1[c(12, 14)] #

matrix1[c(12:14)] #

matrix1[2:4, 1:3] #specify start end points of submatrixes aka tiny matrixes 

#Create a vector of 6 values
values = c(1, 26, 24, 68, 3, 44)
#Create a vector of row names
rnames <- c("R1", "R2") 
#Create a vector of column names
cnames <- c("C1", "C2", "C3") 

matrix4 <- matrix(
  data = values, 
  nrow = 2, 
  ncol = 3, 
  byrow = TRUE,
  dimnames = list(
    rnames, 
    cnames)
  ) # 

matrix4 #

matrix4['R2','C1']
```

### Exercise 1b: Matrices
Compute the mean of column "C2" in matrix4. (Should be 14.5)
```{r Matrix Exercise}
mean(matrix4[,'C2']) #add , because we want columns

```



## Lists
### Creating Lists
```{r Creating Lists}
# Lists ---- can have a mix of objects, numbers, matrixes, etc.
# Creating Lists
list1 <- list(
  name = "Maria", 
  mynumbers = vector1, 
  mymatrix = matrix1, 
  age = 5.3)

```


### List Operations
```{r List Operations}
# List Operations

#
list1

#$ operator!! if you know name of object you want to access use $
list1$age

#
names(list1)

#returns first two elements of list so names and numbers
list1[1:2]

#renaming 
list1$name = "Janelle"

#value of 4th element than use double brackets??
list1[[4]] = 44
```

## Data Frames
### Creating dataframes
```{r Creating data frames}
# Data Frames ---- collections of vectors that can have different types
#Create three vectors of equal length
d <- c(1, 2, 3, 4) # What type of vector? numbers
e <- c("red", "white", "red", NA) # What type of vector?characters
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? logics

#Combine the vectors into a dataframe
dataframe1 <- data.frame(d,e,f) 

#Set the column names & View
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) #opened new window because of View function - make sure to comment view out before turning into PDF

#Add a new column to the dataframe
dataframe1$month <- c('Jan','Feb', 'Mar', 'Apr')

```


### Dataframe operations
```{r Dataframe Operations}
#Extract a column from the dataframe f
dataframe1$Color

#Extract a row
dataframe1[2,]

#Extract several rows
dataframe1[1:2,] 

#Extract a column
dataframe1[,1]

#Extract a column/columns from their name
out1 <- dataframe1['Color']; class(out1)

out2 <- dataframe1$Color; class(out2)

out3 <- dataframe1[c("ID","Passed")]; class(out3)

```